How to select predictive models for decision-making or causal inference?

### 2023-11-19 how_to_select_causal_models.tex
iter condense

There is a rich causal-inference statistical literature, which bridges to
predictive modeling via so-called \emph{outcome models}, or G-computation,
G-formula \cite{robins_role_1986}, Q-model \cite{snowden_implementation_2011},
conditional mean regression \cite{wendling_comparing_2018}. A central challenge
of inference of treatment effects is that of confounding: spurious associations
between treatment allocation and baseline health, \emph{eg} only prescribing a
drug to mild cases \citep{hernan_causal_2020,vanderweele2019principles}.
% Controlled allocation of treatment, as in randomized controlled trials (RCTs),
% alleviate this concern. Yet most machine-learning models are trained on
% \emph{observational} data, close to real-word practice
% \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
Causal inference has been central to epidemiology, typically with methods that
model treatment assignment \cite{austin_moving_2015,grose_use_2020}, based on
propensity scores \cite{rosenbaum_central_1983}.  However, outcome modeling go
beyond average effects, estimating individualized or conditional average
treatment effects (CATE), central to precision medicine.
%
For this purpose, such methods are also invaluable on randomized trials
\cite{su2018random,lamont2018identification,hoogland2021tutorial}.

@@

%\idea{There is a variety of models for g-estimation}
% There is an explosion of outcome modeling or machine learning methods, either 
% developed for
% For instance, many deep-learning methods have been developed for medical
% image analysis \cite{shen2017deep,monshi2020deep}.
Outcome-modeling methods, even when specifically designed for causal
inference are numerous: Bayesian Additive Regression Trees
\cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
\cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
\cite{powers_methods_2018}, causal multivariate adaptive regression
splines \cite{powers_methods_2018}, random forests
\cite{wager_estimation_2018, athey_generalized_2019},
Meta-learners \cite{kunzel_metalearners_2019}, R-learners
\cite{nie_quasioracle_2017}, Doubly robust estimation
\cite{chernozhukov_double_2018}...
%\idea{The variety of estimators calls for model selection procedures}
The wide variety of methods raises the problem
of selecting between different estimators based on the data at hand.
%
Indeed, estimates of treatment effects can vary markedly across different
predictive models (as shown by \cite{dorie_automated_2019} and developed in
Appendix \ref{apd:toy_example:acic_2016_ate_variability}).

@@

Using a predictor to reason on causal effects relies on contrasting the
prediction of the outcome for a given individual with and without the
treatment --as detailed in \autoref{sec:neyman_rubin}.
%
Given various predictors of the outcome, which one should we use?
%
Standard predictive modeling or machine-learning practice selects the predictor
that minimizes the expected error. However, this predictor may not be the best
model to reason about causal effects of an intervention as Figure
\ref{fig:toy_example} illustrates. Consider the probability $Y$ of an undesirable outcome
(\emph{eg} death), a binary treatment $A \in \{0, 1\}$, and a covariate $X \in
    \mathbb R$ summarizing the patient health status (eg. the Charlson index
\cite{charlson_new_1987}). We simulate a treatment beneficial (decreases $Y$)
for patients with high Charlson scores (bad health status) but with little
effect for patients in good condition (low Charlson scores).

@@

\paragraph{Estimating treatment effects with outcome models}\label{subsec:estimators}

Should we know the two expected outcomes for a given $X$,
we could compute the
difference between them, which gives the causal effect of the treatment.
%
These two expected outcomes can be computed from the observed data:
the consistency \ref{assumption:consistency} and unconfoundedness
\ref{assumption:ignorability} assumptions imply the equality of two different
expectations:
\begin{equation}\label{eq:mu_identification}
    \mathbb E_{Y(a) \sim \mathcal{D^{\star}}} [Y(a)|X=x] = \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
\end{equation}
On the left, the expectation is taken on the counterfactual unobserved
distribution. On the right, the expectation is taken on the factual observed
distribution conditionally on the treatment. This equality is referred as the
g-formula identification \cite{robins_new_1986}. For the rest of the
paper, the expectations will always be taken on the factual observed
distribution $\mathcal{D}$. This identification leads to outcome based estimators (ie.
g-computation estimators \cite{snowden_implementation_2011}), targeting the
ATE $\tau$ with outcome modeling:
\begin{eqnarray}
    \tau =& \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)|X=x]
    \notag
    \\
    = &\mathbb E_{Y \sim \mathcal{D}}[Y|A=1] - \mathbb E_{Y \sim \mathcal{D}}[Y| A=0]
    \label{eq:tau_population}
\end{eqnarray}
This equation builds on two quantities: the conditional expectancy
of the outcome given the covariates and either
treatment or no no treatment, called \emph{response function}:
\begin{flalign*}
    \text{Response function}
     &  &
    \mu_{a}(x) \myeq \; \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
     &  &
\end{flalign*}

@@

As extreme inverse propensity weights induce high variance, clipping can be
useful for numerical stability
\cite{swaminathan_counterfactual_2015, ionides_truncated_2008}.

@@

\section{Theory: Links between feasible and oracle risks}\label{sec:theory}

We now relate two feasible risks, $\mu \text{-risk}_{IPW}$ and the
$R\text{-risk}$ to the oracle $\tau\text{-risk}$. Both results make
explicit the role of overlap for the performances of causal risks.

These bounds depend on a specific form of residual that we now define: for each potential outcome, $a \in  \{0; 1\}$, the variance conditionally on $x$
is \cite{shalit_estimating_2017}:
\begin{equation*}\label{eq:residuals}
    \sigma_{y}^{2}(x ; a) \overset{\text{def}}{=}
    \int_{y}\left(y-\mu_{a}(x)\right)^{2} p(y \mid x=x ; A=a) \, d y
\end{equation*}
Integrating over the population, we get the Bayes squared error:
$\sigma^2_{B}(a) = \int_{\mathcal X} \sigma_y^2(x;a) p(x)dx$
and its propensity weighted version:
$\tilde{\sigma}^2_{B}(a) = \int_{\mathcal X}\sigma_y^2(x;a)\,  p(x;
    a)\,dx$. In case of a purely deterministic link between the
covariates, the treatment, and the outcome, these residual terms are null.

\subsection{Upper bound of $\tau\text{-risk}$ with
    $\mu\text{-risk}_{IPW}$}\label{theory:mu_risk_ipw_bound}

\begin{proposition}[Upper bound with $\mu \text{-risk}_{IPW}$
    ]\label{theory:prop:mu_risk_ipw_bound}
    \cite{johansson2022generalization} Given an outcome model $f$, let a
    weighting function\\ $w(x; a) = \frac{a}{e(x)} + \frac{1-a}{1-e(x)}$ as the
    Inverse Propensity Weight. Then, under overlap (assumption
    \ref{assumption:overlap}), we have:
    \begin{equation*}
        \tau\text{-risk}(f) \leq \; 2 \, \mu\text{-risk}_{IPW}(w, f)
        - 2 \, \big(\sigma^2_{B}(1) +  \sigma^2_{B}(0)\big)
    \end{equation*}
\end{proposition}
This result has been derived in previous work
\cite{johansson2022generalization}. It links $\mu\text{-risk}_{IPW}$ to
the squared residuals of each population. For completeness, we provide the proof in \ref{apd:proofs}.

The upper-bound comes from the triangular inequality applied to the residuals of
both populations. The two quantities are equal when the
absolute residuals on treated and untreated populations are equal on the
whole covariate space:
$\forall x \in \mathcal X, |\mu_1(x) - f(x, 1)| = |\mu_0(x) - f(x, 0)|$.
The main difference between the oracle $\tau \text{-risk}$ and the
reweighted mean squared error, $\mu\text{-risk}_{IPW}$, comes from heterogeneous
residuals between populations.
% These quantities are difficult to characterize as
%they are linked both to the estimator and to the data distribution.
%
This bound shows that minimizing the $\mu\text{-risk}_{IPW}$ helps to
minimize the $\tau\text{-risk}$, which leads to
interesting optimization procedures \cite{johansson2022generalization}. However, there is no
guarantee that this bound is tight, which makes it fragile for model
selection.

Assuming strict overlap (probability of all individuals being treated or not
bounded away from 0 and 1 by $\eta$, \ref{apd:causal_assumptions}), the
above bound simplifies into a looser one involving the usual mean squared error:
$\tau\text{-risk}(f)\leq \frac{2}{\eta}\, \mu\text{-risk}(f) -  2 \, \big(\sigma^2_{B}(1) +  \sigma^2_{B}(0)\big)$. For weak overlap (propensity scores not bounded far from 0
or 1), this bound is very loose (as shown in Figure \ref{fig:toy_example})
and is not appropriate to discriminate between models with close performances.

@@

\subsection{Reformulation of the $R\text{-risk}$ as reweighted
    $\tau\text{-risk}$}\label{theory:r_risk_rewrite}

@@

The $R \text{-risk}$ targets the oracle at the cost of an overlap re-weighting
and the addition of the reweighted Bayes residuals, which are independent of
$f$. In good overlap regions the weights $e(x) \big(1-e(x) \big)$ are close to
$\frac{1}{4}$, hence the $R \text{-risk}$ is close to the desired gold-standard
$\tau \text{-risk}$. On the contrary, for units with extreme overlap violation,
these weights go down to zero with the propensity score.

% \begin{remark}[Trivial upper bound] Because $e(x)\leq1$ and $(1-e(x))\leq1$,
%   we see immediately that \begin{equation} R \text{-risk}^* \leqslant
%   \tau{\text{-risk}}+\tilde{\sigma_{B}}(1)^{2}+\tilde{\sigma_{B}}^{2}(0)
%   \end{equation} \end{remark}

% \begin{remark}[Lower bound] Due to strict overlap assumption, \begin{equation}
%   R \text{-risk}^* \geqslant \eta (1 - \eta)
%   \tau{\text{-risk}}-\tilde{\sigma_{B}}(1)^{2}-\tilde{\sigma_{B}}^{2}(0)
%   \end{equation} \end{remark}

% \begin{remark}[Cases of loose upper bounds] Giving a strict overlap assumption
%   $\eta$, we can exhibit for all $C \; st. \; \eta^{-1}>C>1$, simple
%   simulations where: \begin{equation} R \text{-risk}^* \leqslant C \; \eta \;
%   \tau \text{-risk}(f) \quad \end{equation} \end{remark}

\subsection{Interesting special cases}

\paragraph{Randomization special case}\label{remark:rct} If the treatment is
randomized as in RCTs, $p(A=1 \mid X=x) = p(A=1)=p_A$, thus
$\mu\text{-risk}_{IPW}$ takes a simpler form:
\begin{equation*}
    \mu\text{-risk}_{IPW} = \mathbb{E}_{(Y, X, A) \sim \mathcal D}\left[ \Big( \frac{A}{p_A} + \frac{1-A}{1-p_A} \Big) (Y-f(X ; A))^2 \right]
\end{equation*}
However, we still can have large differences
between $\tau\text{-risk}$ and $\mu\text{-risk}_{IPW}$ coming from heterogeneous
errors between populations as noted by Proposition \ref{theory:prop:mu_risk_ipw_bound}
and shown experimentally in \citet{schuler_comparison_2018} and our
results below.

@@

We generate the response functions using random bases. Basis extension methods
are common in biostatistics, \emph{eg} functional regression with splines
\cite{howe_splines_2011, perperoglou_review_2019}. By allowing the function to
vary at specific knots, they give flexible non-linear models. The number of
knots controls the complexity of the function. We use random approximation of
Radial Basis Function (RBF) kernels \cite{rahimi_random_2008} to generate the
outcome and treatment functions. RBF use the same process as polynomial splines
but replace polynomial by Gaussian kernels. Unlike polynomials, Gaussian kernels
have decreasing influences in the input space. This avoids unrealistic
divergences of the response surfaces at the ends of the feature space. We
generate 1\,000 datasets based on this functions, with random overlap
parameters. Example shown in Figure \ref{fig:simulation_examples} and details in
\ref{apd:experiments:generation}.

@@

We compare different candidate estimators that approximate imperfectly the
data-generating process. To build such estimators, we first use a RBF
expansion similar to the one used for data generation. We choose two random
knots and apply a transformation of the raw data features with a Gaussian
kernel. This step is referred as the featurization. Then, we fit a linear
regression on these transformed features. We consider two ways of combining
these steps for outcome mode; we use common nomenclature
\cite{kunzel_metalearners_2019,shen2023rctrep} to refer to these different
meta-learners that differ on how they model, jointly or not, the treated and the
non treated:
\begin{itemize}
    \item SLearner: A single learner for both population, taking the treatment as
          a supplementary covariate.
    \item SftLearner: A single set of basis functions is sampled at random for both
          populations, leading to a given feature space used to model both the treat and
          the non treated, then two
          separate different regressors are fitted on this shared representation.
    \item TLearner: Two completely different learners for each population, hence
          separate feature representations and regressors.
\end{itemize}

We do not include more elaborated meta-learners such as R-learner
\cite{nie_quasioracle_2017} or X-learner
\cite{kunzel_metalearners_2019}. Our goal is not to have the best possible
learner but to have a variety of sub-optimal learners to compare the
different causal metrics. For the same reason, we did not include more powerful
outcome models such as random forests or boosting trees.

@@

Each metric ranks differently the candidate models. Figure
\ref{fig:relative_kendalls_all_datasets} shows the agreement between the
ideal ranking of methods given the oracle $\tau\text{-risk}$ and
the different feasible causal metrics. We measure this agreement
with a relative\footnote{To remove the variance
    across datasets (some datasets lead to easier model selection than
    others), we report values for one metric relative to the mean of all
    metrics for a given dataset instance: $\text{Relative} \, \kappa(\ell,\tau\mathrm{{-risk}})=
        \kappa(\ell,\tau\mathrm{{-risk}}) -
        mean_{\ell}\big(\kappa(\ell,\tau\mathrm{{-risk}})\big)$} Kendall tau $\kappa$ (eq.
\ref{eq:kendall_tau}) \cite{kendall_new_1938}.
Given the importance of overlap in
how well metrics approximate the oracle $\tau\text{-risk}$
(\ref{apd:proofs:mu_risk_ipw_bound}), we separate
strong and weak overlap.

@@

\subsection{Upper bound of $\tau\text{-risk}$ with
    $\mu\text{-risk}_{IPW}$}\label{apd:proofs:mu_risk_ipw_bound}

For the bound with the $\mu\text{-risk}_{IPW}$, we will decompose the CATE risk
on each factual population risks:

\begin{definition}[Population Factual $\mu\text{-risk}$]\label{mu_risk_a}
    \cite{shalit_estimating_2017}
    \begin{equation*}
        \mu\text{-risk}_{a}(f)= \int_{\mathcal Y \times \mathcal X} (y-f(x ; A=a))^{2}  p(y ; x=x \mid A=a) \; dy dx
    \end{equation*}
\end{definition}

Applying Bayes rule, we can decompose the $\mu\text{-risk}$ on each
intervention:
\begin{equation*}
    \mu\text{-risk}(f)
    =p_{A} \,\mu\text{-risk}_{1}(f)+\left(1-p_{A}\right) \,\mu\text{-risk}_{0}(f)
    \text{with } p_A=\mathbb P(A=1)
\end{equation*}

These definitions allows to state a intermediary result on each population:
\begin{lemma}[Mean-variance decomposition]\label{apd:proofs:mu_risk_ipw_link_mu}
    We need a reweighted version of the classical mean-variance decomposition.

    For an outcome model $f: x \times A \rightarrow \mathcal X$. Let the inverse
    propensity weighting function $w(a ; x)=a e(x)^{-1}+(1-a)(1-e(x))^{-1}$.
    \begin{align*}
         & \int_{\mathcal X}(\mu_{1}(x)-f(x ; 1))^{2} p(x) dx  = p_{A} \mu\text{-risk}_{IPW, 1}(w, f)  -\sigma^{2}_{Bayes}(1)
    \end{align*}
    And
    \begin{align*}
         & \int_{\mathcal X}(\mu_{0}(x)-f(x; 0))^{2} p(x) dx   = (1-p_A) \mu\text{-risk}_{IPW, 0}(w, f)  -\sigma^{2}_{Bayes}(0)
    \end{align*}

    \begin{proof}
        \begin{align*}
             & p_{A} \mu\text{-risk}_{IPW, 1}(w, f)  = \int_{\mathcal X \times \mathcal Y} \frac{1}{e(x)}(y-f(x ; 1))^{2} p(y \mid x ; A=1) p(x ; A=1) d y d x                                                      \\
             & = \int_{\mathcal X \times \mathcal Y} (y-f(x ; 1))^{2} p(y \mid x ; A=1) \frac{p(x ; A=1)}{p(x ; A=1)}p(x)dy dx                                                                                      \\
             & = \int_{\mathcal X \times \mathcal Y} \big[(y-\mu_1(x))^{2}+\left(\mu_{1}(x)-f(x ; 1)\right)^{2} + 2\left(y-\mu_{1}(x)\right)\left(\mu_{1}(x)-f(x, 1)\right) \big] p(y \mid x ; A=1) p(x) d y d x    \\
             & =\int_{\mathcal X} \big [ \int_{\mathcal Y} (y-\mu_1(x))^{2} p(y \mid x ; A=1) dy\big ] p(x)dx + \int_{\mathcal X \times \mathcal Y} \left(\mu_{1}(x)-f(x ; 1)\right)^{2} p(x)p(y \mid x ; A=1)dx dy \\
             & + \qquad 2 \int_{\mathcal X} \big [ \int_{\mathcal Y} \left(y-\mu_{1}(x)\right) p(y \mid x ; A=1) dy \big ] \left(\mu_{1}(x)-f(x, 1)\right)p(x)dx                                                    \\
             & =\int_{\mathcal X} \sigma_{y}^{2}(x, 1) p(x) d x +\int_{\mathcal X} \left(\mu_{1}(x)-f(x ; 1)\right)^{2} p(x) d x+0
        \end{align*}
    \end{proof}
\end{lemma}

\begin{proposition*}[Upper bound with mu-IPW]\label{apd:proofs:prop:upper_bound}
    Let f be a given outcome model, let the weighting function $w$ be the Inverse
    Propensity Weight $w(x; a) = \frac{a}{e(x)} + \frac{1-a}{1-e(x)}$. Then, under
    overlap (assumption \ref{assumption:overlap}),
    \begin{align*}
        \tau\text{-risk}(f) \leq & \; 2 \, \mu\text{-risk}_{IPW}(w, f)  \; - 2 \, (\sigma^2_{Bayes}(1) +  \sigma^2_{Bayes}(0))
    \end{align*}

    \begin{proof}
        \begin{align*}
             & \tau\text{-risk}(f) =\int_{\mathcal X}(\mu_{1}(x)-\mu_{0}(x)-(f(x ; 1)-f(x ; 0))^{2} p(x) d x
        \end{align*}
        By the triangle inequality $(u+v)^2 \leq 2(u^2 + v^2)$:
        \begin{align*}
             & \tau\text{-risk}(f) \leq
            2 \int_{\mathcal X}\big[\left(\mu_{1}(x)-f(x ; 1)\right)^{2}+\left(\mu_{0}(x)-f(x ; 0)\right)^{2}\big] p(x) d x \\
        \end{align*}
        Applying Lemma \ref{apd:proofs:mu_risk_ipw_link_mu},
        \begin{align*}
             & \tau\text{-risk}(f) \leq 2\big[p_A \mu\text{-risk}_{IPW, 1}(w, f) +(1-p_A) \mu\text{-risk}_{IPW, 0}(w, f)(w, f)\big] -2(\sigma^{2}_{Bayes}(0) + \sigma^{2}_{Bayes}(1)) \\
             & = 2 \mu\text{-risk}_{IPW}(w, f)-2(\sigma^{2}_{Bayes}(0) + \sigma^{2}_{Bayes}(1))
        \end{align*}
    \end{proof}
\end{proposition*}

@@

>>>>

Predictive modeling bridges to causal-inference with the so-called \emph{outcome
    models}, or G-computation, G-formula \cite{robins_role_1986}, Q-model
\cite{snowden_implementation_2011}, conditional mean regression
\cite{wendling_comparing_2018}. A central challenge of inference of treatment
effects is that of confounding: spurious associations between treatment
allocation and baseline health, \emph{eg} only prescribing a drug to mild cases
\citep{hernan_causal_2020,vanderweele2019principles}.
% Controlled allocation of treatment, as in randomized controlled trials (RCTs),
% alleviate this concern. Yet most machine-learning models are trained on
% \emph{observational} data, close to real-word practice
% \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
Causal inference has been central to epidemiology, typically with methods that
model treatment assignment \cite{austin_moving_2015,grose_use_2020}, based on
propensity scores \cite{rosenbaum_central_1983}. But outcome modeling go
beyond average effects, estimating individualized or conditional average
treatment effects (CATE), central to precision medicine.
%
For this purpose, such methods are also invaluable on randomized trials
\cite{su2018random,lamont2018identification,hoogland2021tutorial}.

@@

%\idea{There is a variety of models for g-estimation}
% There is an explosion of outcome modeling or machine learning methods, either 
% developed for
% For instance, many deep-learning methods have been developed for medical
% image analysis \cite{shen2017deep,monshi2020deep}.
Outcome-modeling methods, even when specifically designed for causal
inference are numerous: Bayesian Additive Regression Trees
\cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
\cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
\cite{powers_methods_2018}, causal multivariate adaptive regression
splines \cite{powers_methods_2018}, random forests
\cite{wager_estimation_2018, athey_generalized_2019},
Meta-learners \cite{kunzel_metalearners_2019}, R-learners
\cite{nie_quasioracle_2017}, Doubly robust estimation
\cite{chernozhukov_double_2018}...
%\idea{The variety of estimators calls for model selection procedures}
The wide variety of methods raises the problem
of selecting between different estimators based on the data at hand.%TODO: might be removed since repeated two sentences below
%
Indeed, estimates of treatment effects can vary markedly across different
predictive models (as shown by \cite{dorie_automated_2019} and developed in
Appendix \ref{apd:toy_example:acic_2016_ate_variability}).

@@

Using a predictor to reason on causal effects relies on contrasting the
prediction of the outcome for a given individual with and without the treatment.
%
Given various predictors of the outcome, which one should we use?
%
Standard predictive modeling or machine-learning practice selects the predictor
that minimizes the expected error. However, this predictor may not be the best
model to reason about causal effects of an intervention as Figure
\ref{fig:toy_example} illustrates. Consider the probability $Y$ of an undesirable outcome
(\emph{eg} death), a binary treatment $A \in \{0, 1\}$, and a covariate $X \in
    \mathbb R$ summarizing the patient health status (eg. the Charlson index
\cite{charlson_new_1987}). We simulate a treatment beneficial (decreases mortality)
for patients with high Charlson scores (bad health status) but with little
effect for patients in good condition (low Charlson scores).

@@

\paragraph{Estimating treatment effects with outcome models -- g-computation \cite{robins_new_1986}}\label{subsec:estimators}

Should we know the two expected outcomes for a given $X$,
we could compute the
difference between them, which gives the causal effect of the treatment.
%
These two expected outcomes can be computed from the observed data:
the consistency \ref{assumption:consistency} and unconfoundedness
\ref{assumption:ignorability} assumptions imply the equality of two different
expectations:
\begin{equation}\label{eq:mu_identification}
    \mathbb E_{Y(a) \sim \mathcal{D^{\star}}} [Y(a)|X=x] = \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
\end{equation}
On the left, the expectation is taken on the counterfactual unobserved
distribution. On the right, the expectation is taken on the factual observed
distribution conditionally on the treatment. For the rest of the
paper, the expectations will always be taken on the factual observed
distribution $\mathcal{D}$. This identification leads to outcome based estimators (ie.
g-computation estimators \cite{snowden_implementation_2011}):
\begin{eqnarray}
    \tau =& \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)|X=x]
    \notag
    \\
    = &\mathbb E_{Y \sim \mathcal{D}}[Y|A=1] - \mathbb E_{Y \sim \mathcal{D}}[Y| A=0]
    \label{eq:tau_population}
\end{eqnarray}
This equation builds on two quantities: the conditional expectancy
of the outcome given the covariates and either
treatment or no no treatment, called \emph{response function}:
\begin{flalign*}
    \text{Response function}
     &  &
    \mu_{a}(x) \myeq \; \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
     &  &
\end{flalign*}

@@

% As extreme inverse propensity weights induce high variance, clipping can be
% useful for numerical stability
% \cite{swaminathan_counterfactual_2015, ionides_truncated_2008}.

@@

\section{A theoretical argument favoring the R-risk}\label{theory:r_risk_rewrite}

% We now one feasible risks, $\mu \text{-risk}_{IPW}$ and the
% $R\text{-risk}$ to the oracle $\tau\text{-risk}$. Both results make
% explicit the role of overlap for the performances of causal risks.

This rewriting involves reweighted residuals: for each potential outcome, $a \in
    \{0; 1\}$, the variance conditionally on $x$ is \cite{shalit_estimating_2017}:
\begin{equation*}\label{eq:residuals}
    \sigma_{y}^{2}(x ; a) \overset{\text{def}}{=}
    \int_{y}\left(y-\mu_{a}(x)\right)^{2} p(y \mid x=x ; A=a) \, d y
\end{equation*}
Integrating over the population, we get the Bayes squared error:
$\sigma^2_{B}(a) = \int_{\mathcal X} \sigma_y^2(x;a) p(x)dx$
and its propensity weighted version:
$\tilde{\sigma}^2_{B}(a) = \int_{\mathcal X}\sigma_y^2(x;a)\,  p(x;
    a)\,dx$. In case of a purely deterministic link between the
covariates, the treatment, and the outcome, these residual terms are null.

@@

@@

The $R \text{-risk}$ targets the oracle at the cost of an overlap re-weighting
and the addition of the reweighted Bayes residuals, which are independent of
$f$. In good overlap regions the weights $e(x) \big(1-e(x) \big)$ are close to
$\frac{1}{4}$, hence the $R \text{-risk}$ is close to the desired gold-standard
$\tau \text{-risk}$. For randomized control trials, this weight is constant
making the $R\text{-risk}$ particularly suited for exploring
heterogeneity (Details in Appendix \ref{apd:theory:special_cases})

@@

We generate the response functions using random bases extension, a common method in
biostatistics, \emph{eg} functional regression with splines
\cite{howe_splines_2011, perperoglou_review_2019}. By allowing the function to
vary at specific knots, they give flexible non-linear models. The number of
knots controls the complexity of the function. We use random approximation of
Radial Basis Function (RBF) kernels \cite{rahimi_random_2008} to generate the
outcome and treatment functions. RBF use the same process as polynomial splines
but replace polynomial by Gaussian kernels. Unlike polynomials, Gaussian kernels
have decreasing influences in the input space. This avoids unrealistic
divergences of the response surfaces at the ends of the feature space. We
generate 1\,000 datasets based on these functions, with random overlap
parameters. Example shown in Figure \ref{fig:simulation_examples} and details in
\ref{apd:experiments:generation}.

@@

We compare different candidate estimators that approximate imperfectly the
data-generating process. To build such estimators, we first use a RBF
expansion similar to the one used for data generation. We choose two random
knots and apply a transformation of the raw data features with a Gaussian
kernel. This step is referred as the featurization. Then, we fit a linear
regression on these transformed features. We consider two ways of combining
these steps for outcome mode; we use common nomenclature
\cite{kunzel_metalearners_2019,shen2023rctrep} to refer to these different
meta-learners that differ on how they model, jointly or not, the treated and the
non treated:
\begin{itemize}
    \item SLearner: A single learner for both population, taking the treatment as
          a supplementary covariate.
    \item SftLearner: A single set of basis functions is sampled at random for both
          populations, leading to a given feature space used to model both the treated and
          controls, then two
          separate different regressors are fitted on this shared representation.
    \item TLearner: Two completely different learners for each population, hence
          separate feature representations and regressors.
\end{itemize}

% We do not include more elaborated meta-learners such as R-learner
% \cite{nie_quasioracle_2017} or X-learner
% \cite{kunzel_metalearners_2019}. Our goal is not to have the best possible
% learner but to have a variety of sub-optimal learners to compare the
% different causal metrics. For the same reason, we did not include more powerful
% outcome models such as random forests or boosting trees.

@@

Figure \ref{fig:relative_kendalls_all_datasets} shows the agreement between the
ideal ranking of outcome models given the oracle $\tau\text{-risk}$ and the different
feasible causal metrics. We measure this agreement with a relative\footnote{To
    remove the variance across datasets (some datasets lead to easier model
    selection than others), we report values for one metric relative to the mean of
    all metrics for a given dataset instance: $\text{Relative} \,
        \kappa(\ell,\tau\mathrm{{-risk}})= \kappa(\ell,\tau\mathrm{{-risk}}) -
        mean_{\ell}\big(\kappa(\ell,\tau\mathrm{{-risk}})\big)$} Kendall tau $\kappa$
(eq. \ref{eq:kendall_tau}) \cite{kendall_new_1938}. Given the importance of
overlap in how well metrics approximate the oracle $\tau\text{-risk}$, we
separate strong and weak overlap.

@@

% \subsection{Upper bound of $\tau\text{-risk}$ with
%     $\mu\text{-risk}_{IPW}$}\label{apd:proofs:mu_risk_ipw_bound}

% For the bound with the $\mu\text{-risk}_{IPW}$, we will decompose the CATE risk
% on each factual population risks:

% \begin{definition}[Population Factual $\mu\text{-risk}$]\label{mu_risk_a}
%     \cite{shalit_estimating_2017}
%     \begin{equation*}
%         \mu\text{-risk}_{a}(f)= \int_{\mathcal Y \times \mathcal X} (y-f(x ; A=a))^{2}  p(y ; x=x \mid A=a) \; dy dx
%     \end{equation*}
% \end{definition}

% Applying Bayes rule, we can decompose the $\mu\text{-risk}$ on each
% intervention:
% \begin{equation*}
%     \mu\text{-risk}(f)
%     =p_{A} \,\mu\text{-risk}_{1}(f)+\left(1-p_{A}\right) \,\mu\text{-risk}_{0}(f)
%     \text{with } p_A=\mathbb P(A=1)
% \end{equation*}

% These definitions allows to state a intermediary result on each population:
% \begin{lemma}[Mean-variance decomposition]\label{apd:proofs:mu_risk_ipw_link_mu}
%     We need a reweighted version of the classical mean-variance decomposition.

%     For an outcome model $f: x \times A \rightarrow \mathcal X$. Let the inverse
%     propensity weighting function $w(a ; x)=a e(x)^{-1}+(1-a)(1-e(x))^{-1}$.
%     \begin{align*}
%          & \int_{\mathcal X}(\mu_{1}(x)-f(x ; 1))^{2} p(x) dx  = p_{A} \mu\text{-risk}_{IPW, 1}(w, f)  -\sigma^{2}_{Bayes}(1)
%     \end{align*}
%     And
%     \begin{align*}
%          & \int_{\mathcal X}(\mu_{0}(x)-f(x; 0))^{2} p(x) dx   = (1-p_A) \mu\text{-risk}_{IPW, 0}(w, f)  -\sigma^{2}_{Bayes}(0)
%     \end{align*}

%     \begin{proof}
%         \begin{align*}
%              & p_{A} \mu\text{-risk}_{IPW, 1}(w, f)  = \int_{\mathcal X \times \mathcal Y} \frac{1}{e(x)}(y-f(x ; 1))^{2} p(y \mid x ; A=1) p(x ; A=1) d y d x                                                      \\
%              & = \int_{\mathcal X \times \mathcal Y} (y-f(x ; 1))^{2} p(y \mid x ; A=1) \frac{p(x ; A=1)}{p(x ; A=1)}p(x)dy dx                                                                                      \\
%              & = \int_{\mathcal X \times \mathcal Y} \big[(y-\mu_1(x))^{2}+\left(\mu_{1}(x)-f(x ; 1)\right)^{2} + 2\left(y-\mu_{1}(x)\right)\left(\mu_{1}(x)-f(x, 1)\right) \big] p(y \mid x ; A=1) p(x) d y d x    \\
%              & =\int_{\mathcal X} \big [ \int_{\mathcal Y} (y-\mu_1(x))^{2} p(y \mid x ; A=1) dy\big ] p(x)dx + \int_{\mathcal X \times \mathcal Y} \left(\mu_{1}(x)-f(x ; 1)\right)^{2} p(x)p(y \mid x ; A=1)dx dy \\
%              & + \qquad 2 \int_{\mathcal X} \big [ \int_{\mathcal Y} \left(y-\mu_{1}(x)\right) p(y \mid x ; A=1) dy \big ] \left(\mu_{1}(x)-f(x, 1)\right)p(x)dx                                                    \\
%              & =\int_{\mathcal X} \sigma_{y}^{2}(x, 1) p(x) d x +\int_{\mathcal X} \left(\mu_{1}(x)-f(x ; 1)\right)^{2} p(x) d x+0
%         \end{align*}
%     \end{proof}
% \end{lemma}

% \begin{proposition*}[Upper bound with mu-IPW]\label{apd:proofs:prop:upper_bound}
%     Let f be a given outcome model, let the weighting function $w$ be the Inverse
%     Propensity Weight $w(x; a) = \frac{a}{e(x)} + \frac{1-a}{1-e(x)}$. Then, under
%     overlap (assumption \ref{assumption:overlap}),
%     \begin{align*}
%         \tau\text{-risk}(f) \leq & \; 2 \, \mu\text{-risk}_{IPW}(w, f)  \; - 2 \, (\sigma^2_{Bayes}(1) +  \sigma^2_{Bayes}(0))
%     \end{align*}

%     \begin{proof}
%         \begin{align*}
%              & \tau\text{-risk}(f) =\int_{\mathcal X}(\mu_{1}(x)-\mu_{0}(x)-(f(x ; 1)-f(x ; 0))^{2} p(x) d x
%         \end{align*}
%         By the triangle inequality $(u+v)^2 \leq 2(u^2 + v^2)$:
%         \begin{align*}
%              & \tau\text{-risk}(f) \leq
%             2 \int_{\mathcal X}\big[\left(\mu_{1}(x)-f(x ; 1)\right)^{2}+\left(\mu_{0}(x)-f(x ; 0)\right)^{2}\big] p(x) d x \\
%         \end{align*}
%         Applying Lemma \ref{apd:proofs:mu_risk_ipw_link_mu},
%         \begin{align*}
%              & \tau\text{-risk}(f) \leq 2\big[p_A \mu\text{-risk}_{IPW, 1}(w, f) +(1-p_A) \mu\text{-risk}_{IPW, 0}(w, f)(w, f)\big] -2(\sigma^{2}_{Bayes}(0) + \sigma^{2}_{Bayes}(1)) \\
%              & = 2 \mu\text{-risk}_{IPW}(w, f)-2(\sigma^{2}_{Bayes}(0) + \sigma^{2}_{Bayes}(1))
%         \end{align*}
%     \end{proof}
% \end{proposition*}

@@

\subsection{Interesting special cases}\label{apd:theory:special_cases}

\paragraph{Randomization special case}\label{remark:rct} If the treatment is
randomized as in RCTs, $p(A=1 \mid X=x) = p(A=1)=p_A$, thus
$\mu\text{-risk}_{IPW}$ takes a simpler form:
\begin{equation*}
    \mu\text{-risk}_{IPW} = \mathbb{E}_{(Y, X, A) \sim \mathcal D}\left[ \Big( \frac{A}{p_A} + \frac{1-A}{1-p_A} \Big) (Y-f(X ; A))^2 \right]
\end{equation*}
However, we still can have large differences
between $\tau\text{-risk}$ and $\mu\text{-risk}_{IPW}$ coming from heterogeneous
errors between populations as shown experimentally in \citet{schuler_comparison_2018} and our
results below.
